---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "doc/Figures/README-",
                      out.width = "100%")
# for tibbles...
options(pillar.neg=F, # do no print neg number in red
        pillar.subtle=F, # turn off highlighting of significant digits
        tibble.width = 170) # default=95, increase it to make it readable
library(tidyverse)
devtools::build_manual()
file.copy("../RegressionCalibration_0.1.0.pdf", "./doc/RegressionCalibration-manual.pdf")
library(RegressionCalibration)

```

:information\_source: `RegressionCalibration` is still under active development.

## Overview
[//]:*******

`RegressionCalibration` is an R-package to perform three-sample Mendelian Randomisation (MR) analyses using the Regression Calibration framework. MR estimates can be subject to bias due the use of weak instruments and winner’s curse. The regression calibration framework can be used when three samples (discovery and replication for the exposure, and outcome) to account for these biases and re-calibrate the causal effect estimate.           
This package builds up on the [`TwoSampleMR`](https://github.com/MRCIEU/TwoSampleMR/) R-package for the implementation of the most common MR estimators. We also recommend using `TwoSampleMR` functions to pre-process the data (identify independent instruments, harmonise data, etc...).


There are several functions available:  
-   **`RC()`**   
main function, to perform regression calibration for all estimators (IVW, simple/weighted median, simple/weighted mode)   
-   **`RC_ivw()`**  
IVW implementation in the regression calibration framework   
-   **`RC_simple_median()`**  
simple median implementation in the regression calibration framework   
-   **`RC_weighted_median()`**  
weigthed median implementation in the regression calibration framework   
-   **`RC_simple_mode()`**  
simple mode implementation in the regression calibration framework   
-   **`RC_weighted_mode()`**  
weigthed mode implementation in the regression calibration framework    

We also implemented the MR-RAPS using ZP and MR-RAPS using UMVCUE approaches:    
-   **`raps_ZP()`**  
-   **`raps_UMVCUE`()`**  

More details can be found in the [manual](doc/RegressionCalibration-manual.pdf).      

## Installation
[//]:*******

You can install the current version of `RegressionCalibration` with:    
```{r install-package, echo=TRUE, eval=F, message=FALSE, results='hide'}
# Directly install the package from github
# install.packages("remotes")
remotes::install_github("n-mounier/RegressionCalibration")
library(RegressionCalibration)
```
<!--- Note: using remotes instead of devtools leads to re-build the package
and apparently, it may be a problem with R 3.4 and macOS, 
see https://stackoverflow.com/questions/43595457/alternate-compiler-for-installing-r-packages-clang-error-unsupported-option/43943631#43943631 --->

## Usage - Main Function
[//]:*******

The `RegressionCalibration` R-Package has been designed to be compatible with the `TwoSampleMR` R-Package: you can use it to extract data and pre-process the data.   
For example, we can use it to obtain BMI data from 3 different samples (UKB, GIANT-males, GIANT-females):

```{r getData, eval=T}

# BMI (exposure - discovery) UKB Neale Lab 2017 
disc <- TwoSampleMR::extract_instruments(outcomes = 'ukb-a-248') 

# BMI (exposure - replication) GIANT 2015 males + BMI (outcome) GIANT 2015 females 
rep_out <- TwoSampleMR::extract_outcome_data(snps = disc$SNP, outcomes =c('ieu-a-785', 'ieu-a-974'))

# harmonise
dat <- TwoSampleMR::harmonise_data(disc, rep_out)

``` 

Note that when extracted data for the replication and for the outcome samples using `TwoSampleMR::extract_outcome_data()`. To distinguish the two, an extra column is needed: 

```{r formatData, eval=T}

# add column for regression calibration: sample should be "replication" or "outcome"
dat %>%
  mutate(sample = case_when(
    id.outcome == 'ieu-a-785' ~ 'replication',
    id.outcome == 'ieu-a-974' ~ 'outcome')) -> dat

``` 

The `dat` object now contains all information needed to perform the Regression Calibration analyses:

```{r analyseData, eval=T}
# columns needed are: 
# SNP, exposure, outcome, beta.exposure, beta.outcome, se.exposure, se.outcome, mr_keep, sample
resRC = RegressionCalibration::RC(dat)
print(resRC)
``` 

The output of the `RegressionCalibration::RC` is similar to the one of `TwoSampleMR::mr()`.    
For this example, we can compare the results using the Regression Calibration approach to the ones obtained without correction for Winner's Curse and weak instrument bias.  

```{r compareResults, eval=T}
# just discovery -> outcome
resMR = TwoSampleMR::mr(dat %>% filter(sample == "outcome"))

# compare IVW estimates
resMR %>% filter(method=="Inverse variance weighted")
resRC %>% filter(method=="Inverse variance weighted")
``` 

The causal effect of BMI on itself is expected to be 1, but we can see that the standard IVW estimate is biased towards the null (`r  resMR %>% filter(method=="Inverse variance weighted") %>% pull(b) %>% round(3)`, SE=`r resMR %>% filter(method=="Inverse variance weighted") %>% pull(se) %>% round(3)`). The IVW using Regression Calibration approach recovers the correct value (`r resRC %>% filter(method=="Inverse variance weighted") %>% pull(b) %>% round(3)`, SE=`r resRC %>% filter(method=="Inverse variance weighted") %>% pull(se) %>% round(3)`).

## Usage - Other Functions

Other functions (method-specific) are also available. As they require a different type of input (vectors of genetic effects and standard errors) they should be use with care.

```{r otherFunctions, eval=T}
dat_disc = dat %>%
  filter(sample=="replication", mr_keep) %>%
  transmute(SNP,
            EA = effect_allele.exposure,
            OA = other_allele.exposure,
            beta = beta.exposure,
            se = se.exposure)
dat_rep = dat %>%
  filter(sample=="replication", mr_keep) %>%
  transmute(SNP,
            EA = effect_allele.outcome,
            OA = other_allele.outcome,
            beta = beta.outcome,
            se = se.outcome)
dat_out = dat %>%
  filter(sample=="outcome", mr_keep) %>%
  transmute(SNP,
            EA = effect_allele.outcome,
            OA = other_allele.outcome,
            beta = beta.outcome,
            se = se.outcome)
# check that for each sample-specific data we have the same SNPs in the same order
table(dat_disc$OA == dat_rep$OA)
table(dat_disc$EA == dat_rep$EA)
table(dat_disc$SNP == dat_out$SNP)
table(dat_disc$OA == dat_out$OA)
table(dat_disc$EA == dat_out$EA)

RegressionCalibration::RC_ivw(b_disc = dat_disc$beta,
                              b_rep = dat_rep$beta,
                              b_out = dat_out$beta,
                              se_disc = dat_disc$se,
                              se_rep = dat_rep$se,
                              se_out = dat_out$se)

# MR-RAPS using ZP correction (as presented in the paper) 
# just discovery -> outcome
RegressionCalibration::raps_ZP(b_exp = dat_disc$beta,
                               b_out = dat_out$beta,
                               se_exp = dat_disc$se,
                               se_out = dat_out$se,
                               threshold = 5e-8)
# less biased than standard IVW, but still not recovering true causal effect

# MR-RAPS using UMVCUE correction (as presented in the paper)
RegressionCalibration::raps_UMVCUE(b_disc = dat_disc$beta,
                               b_rep = dat_rep$beta,
                               b_out = dat_out$beta,
                               se_disc = dat_disc$se,
                               se_rep = dat_rep$se,
                               se_out = dat_out$se,
                               threshold = 5e-8)
# less biased than standard IVW, but still not recovering true causal effect

``` 

## Citation
[//]:*******


## Contact
<mounier.ninon@gmail.com>
